Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-05-15 SCP2008/paper/metamonads.tex
move paper to here

In this paper we report on progress of solving these problems using GE
as our running example. Specifically, our contributions:
\vspace*{-2pt}
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{MSP:PADL04,KiselyovTaha} for generating control structures
      such as loops and conditionals. The extension is non-trivial
      because of control dependencies and because
      let-insertion, as we argue, is a control effect on its own.
\begin{comment}
      : e.g.,\\
      |let x = exp in ...| has a different \emph{effect} within a
      conditional branch.
\end{comment}
    \item Implementation of the |doM|-notation (patterned after
      |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{assure composability of aspects} even
      for aspects that use state and have to be accounted in many
      places in the generated code.
    \item Use functor type sharing constraints to encode domain-specific
      knowledge.
\end{itemize}

@@

We build code generators out of primitive ones using code generation 
combinators. MetaOCaml, as an instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the needed
features: to construct a code expression, to combine them, and to
execute them. Figure~\ref{easycode} shows the simplest code generator |one|,
as well as more complex generators.

\begin{figure}
\begin{code}
let one = .<1>. and plus x y = .<.~x + .~y>.
let simplest_code = let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\evalresult{.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.}
let simplest_param_code plus one = let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
let param_code1 plus one =
  let gen x y = plus (plus y one) (plus x (plus y one)) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
let param_code1' plus one =
  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
param_code1' plus one
\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
\end{code}
\caption{Code generation and combinators. $\Longrightarrow$ under an
  expression shows the result of its evaluation}\label{easycode}
\end{figure}

We use MetaOCaml brackets |.<...>.| to generate code expressions,
i.e., to construct future-stage computations. We use escapes |.~| to
perform an immediate code generating computation \emph{while} we are
building the future-stage computation. The immediate computation in
|simplest_code| is the evaluation of the function |gen|, which in turn
applies |plus|. The function |gen| receives code expressions |.<x>.|
and |.<y>.| as arguments. At the generating stage, we can manipulate
code expressions as (opaque) values. The function |gen| returns a code
expression, which is inlined in the place of the escape. MetaOCaml can
print out code expressions, so we can see the final generated code. It
has no traces of |gen| and |plus|: their applications are done at the
generation stage.

@@

To see the benefit of code generation, we notice that we can easily
parameterize our code, |simplest_param_code|, and use it to generate
code that operates on integers, floating point numbers or booleans --
in general, any domain that implements |plus| and |one|.

The generator |param_code1| has two occurrences of |plus y one|,
which may be quite a complex computation and so we would rather not do
it twice. We may be tempted to rely on the compiler's
common-subexpression elimination optimization. When the generated code is
very complex, however, the compiler may overlook common subexpressions.  Or the
subexpressions may occur in such an imperative context where the compiler
might not be able to determine if lifting them is sound. So, being
conservative, the optimizer will leave the duplicates as they are. We
may attempt to eliminate subexpressions as in |param_code1'|. However,
the result of |param_code1' plus one| still exhibits duplicate
sub-expressions.  Our |let|-insertion optimization saved the
computation at the generating stage.  We need a combinator that
inserts the |let| expression in the generat\emph{ed} code. We need a
combinator |letgen| to be used as\\
|let ce = letgen (plus y one) in plus ce (plus x ce)|
yielding the code like |.<let t = y + 1 in t + (x + t)>.| 
But that seems impossible because |letgen exp| has to generate
the expression |.<let t = exp in body>.| but |letgen| does not
have the |body| yet. The body needs a temporary identifier |.<t>.|
that is supposed to be the result of |letgen| itself.  Certainly
|letgen| cannot generate only part of a let-expression, without the
|body|, as all generated expressions in MetaOCaml are well-formed and
complete.

@@

\noindent Comparison of the code that did let-insertion at the generating stage\\
|let ce = (plus y one) in  plus ce (plus x ce)|\\
with the corresponding code inserting let at the generated code stage\\
|letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))|\\
clearly shows the difference between  direct-style and CPS code.
What was\\ |let ce = init in ...| in direct style became
|init' (fun ce -> ...)| in CPS. For one thing, |let| became
``inverted''. For another, what used to be an expression that yields
a value, |init|, became an expression that takes an extra argument,
the continuation, and invokes it. The differences look negligible in
the above example. In larger expressions with many let-forms, the
number of parentheses around |fun| increases, the need to add and
then invoke the |k| continuation argument become increasingly annoying. The
inconvenience is great enough for some people to explicitly avoid CPS
or claim that numerical programmers (our users) cannot or will not
program in CPS. Clearly a better notation is needed.

@@

let l1 f = fun x -> doM \{ t <-- x; f t\}
let l2 f = fun x y -> doM \{ tx <-- x; ty <-- y; f tx ty\}

@@

Finally, |runM| runs our monad, that is, performs the computation of
the monad and returns its result, which in our case is the code
expression. We run the monad by passing it the initial state and the
initial continuation |k0|. We can now re-write our |param_code2|
example of the previous section as |param_code3|.

\begin{figure}
\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
let param_code4 plus one =
  let gen x y = doM \{ ce <-- retN (plus y one);
                      ret (plus ce (plus x ce)) \}
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
let ifM' test th el = doM \{
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc\}
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 i_2 -> \protect\newline
let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 i_2 -> 
if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
\end{figure}
% param_code3 plus one;;

That does not seem like much of an improvement. With the help of
camlp4 pre-processor, we introduce the |doM|-notation \cite{metamonadsURL},
patterned after the |do|-notation of Haskell. The function
|param_code4|, written in the |doM|-notation, is equivalent to
|param_code3| -- in fact, the camlp4 preprocessor will convert the
former into the latter. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its arguments are
already generated pieces of code rather than monadic values. We
``lift it'', see |ifM'|. We define functions |l1|,
|l2|, |l3| (analogues of |liftM|, |liftM2|, |liftM3| of Haskell) 
to make such a lifting generic. However we also need
another |ifM| function, with the same
interface (see Figure~\ref{ourmonad}). The difference between them is
apparent: in the code above with |ifM'|, the let-insertion
happened \emph{before} the if-expression, that is, before the test that
the index |i| is positive. If |i| turned out
negative, |a.(i)| would generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that the code generation (such as the one in |retN|) is 
truly an effect and we have to be clear about the sequencing of
effects when generating control constructions such as conditionals.
The form |ifM| handles such effects correctly. We
need similar operators for other OCaml control forms: for generating
case-matching statements and |for|- and |while|-loops.

\section{Aspects and Functors}\label{functors}

The monad represents finer-scale code generation. We need tools for
larger-scale modularization; we can use any abstraction
mechanisms we want to structure our code generators, as long as none
of those abstractions infiltrate the generated code.

While the Object-Oriented Design community has acquired an extensive
vocabulary for describing modularity ideas, the guiding principles for
modular designs has not changed since they were first articulated by
Parnas~\cite{journals/cacm/parnas72a} and Dijkstra~
\cite{EWD:EWD447}: information hiding and separation of concerns.  To
apply these principles to the study of Gaussian Elimination, we need
to understand what are the changes between different implementations, and 
what concerns need to be addressed.  We also need to study the degree
to which these concerns are independent.
A study of Gaussian Elimination \cite{carette04} shows that
the following variations occur:
\begin{enumerate}
\vspace*{-6pt}
    \item \textbf{Domain}: In which (algebraic) domain do the
      matrix elements belong to.  Sometimes the domains are very
      specific (e.g., $\mathbb{Z}, \mathbb{Q}, \mathbb{Z}_p$ and
      floating point numbers), while in other cases the domains
      were left generic, e.g., multivariate polynomials over a
      field.  In the roughly 85 pieces of code surveyed
      \cite{carette04} 20 different domains were encountered.
    \item \textbf{Container}: Whether the matrix is represented as
      an array of arrays, a one-dimensional array, a hash table, a
      sparse matrix, etc., and whether indexing is done in C or
      Fortran style.  Additionally, if a particular representation
      had a special mechanism for efficient row exchanges.
    \item \textbf{Output choices}: Whether just the reduced
      matrix, or additionally the rank, the determinant, and the
      pivoting matrix are to be returned. In the larger algorithm
      family, routines like Maple's
      \texttt{LinearAlgebra:-LUDecomposition} have up to $2^6 +
      2^5 + 2^2 = 100$ outputs.
    \item \textbf{Fraction-free}: Whether the Gaussian Elimination
        algorithm is allowed to use unrestricted division, or only
        exact (remainder-free) division.
    \item \textbf{Pivoting}: Whether to use no, 
        column-wise, or full pivoting.
    \item \textbf{Augmented Matrices}: Whether all or only some
      columns of the matrix participate in elimination. We currently
        do not implement this aspect.
\end{enumerate}
\noindent In addition to the above variations, there are two aspects that 
recur frequently:
\vspace*{-6pt}
\begin{enumerate}
    \item \textbf{Length measure}:  For stability reasons
        (numerical or coefficient growth), if a domain
      possesses
        an appropriate length measure, this is sometimes used to choose
        an ``optimal'' pivot.
    \item \textbf{Normalization and zero-equivalence}: Whether the
      arithmetic operations of the domain give normalized results, and
      whether a specialized zero-equivalence routine is to be used.
\end{enumerate}
\noindent These are separated out from the others as they are cross-cutting
concerns: in the case of the length measure, a property of the domain
will influence the pivoting method \emph{if} pivoting is to be
performed.

@@

\begin{code2}
module type OUTPUT = sig
  type contr  type res
  module D : DETERMINANT   module R : RANK   module P : TRACKPIVOT
  val make_result : ('a,contr) code -> 
    (('a,res) code,
     [> `TDet of 'a D.lstate | `TRan of 'a R.lstate | `TPivot of 'a P.lstate]
       list, ('a,'w) code) monad
end
module OutDetRank(Dom:DOMAIN)(C: CONTAINER2D)
    (Det : DETERMINANT with type indet = Dom.v and type outdet = Dom.v)
    (Rank : RANK) = struct
  module Ctr = C(Dom)
  type contr = Ctr.contr
  type res = contr * Det.outdet * int
  module D = Det   module R = Rank   module P = DiscardPivot
  let make_result b = doM \{ det  <-- D.fin ();  rank <-- R.fin ();
    ret .< ( .~b, .~det, .~rank ) >. \}
end
\end{code2}

@@

\begin{figure}
\begin{code2}
module Gen(Dom: DOMAIN)(C: CONTAINER2D)(PivotF: PIVOT)
          (Update: UPDATE with type baseobj = Dom.v and type ctr = C(Dom).contr)
          (Out: OUTPUT with type contr = C(Dom).contr and type D.indet = Dom.v 
                        and type 'a D.lstate = 'a Update.D.lstate) = struct
    module Ctr = C(Dom)
    module Pivot = PivotF(Dom)(C)(Out.D)
    let gen =
      let zerobelow b r c m n brc =
        let innerbody i = doM \{
            bic <-- Ctr.get b i c;
            whenM (l1 LogicCode.not (LogicCode.equal bic Dom.zero ))
                (seqM (retLoopM (Idx.succ c) (Idx.pred m)
                          (fun k -> Update.update b r c i k) )
                      (Ctr.set b i c Dom.zero)) \} in 
        doM \{
              seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
                   (Update.update_det brc) \} in
      let dogen a = doM \{
          r <-- Out.R.decl ();
          c <-- retN (liftRef Idx.zero);
          b <-- retN (Ctr.mapper Dom.normalizerf (Ctr.copy a));
          m <-- retN (Ctr.dim1 a);
          n <-- retN (Ctr.dim2 a);
          () <-- Update.D.decl ();
          () <-- Out.P.decl ();
          seqM 
            (retWhileM (LogicCode.and_ (Idx.less (liftGet c) m)
                                       (Idx.less (liftGet r) n) )
               ( doM \{
               rr <-- retN (liftGet r);
               cc <-- retN (liftGet c);
               pivot <-- l1 retN (Pivot.findpivot b rr m cc n);
               seqM (retMatchM pivot (fun pv -> 
                        seqM (zerobelow b rr cc m n pv)
                             (Out.R.succ ()) )
                        (Update.D.zero_sign () ))
                    (Code.update c Idx.succ) \} ))
            (Out.make_result b) \} in
    .<fun a -> .~(runM (dogen .<a>.)) >.
end
\end{code2}
\end{figure}

@@

\bibliography{metamonads}
\bibliographystyle{plain}

>>>>

In this paper we report on progress of solving these problems using GE
as our running example. Specifically, our contributions:
\vspace*{-2pt}
\begin{itemize}
    \item Extending a let-insertion, memoizing monad of
      \cite{MSP:PADL04,KiselyovTaha} for generating control structures
      such as loops and conditionals. The extension is non-trivial
      because of control dependencies and because
      let-insertion, as we argue, is a control effect on its own.
      : e.g.,\\
      |let x = exp in ...| has a different \emph{effect} within a
      conditional branch.
    \item Implementation of the |perform|-notation (patterned after
      |do|-notation of Haskell) to make monadic code readable.
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{assure composability of aspects} even
      for aspects that use state and have to be accounted in many
      places in the generated code.
    \item Use functor type sharing constraints to encode domain-specific
      knowledge.
\end{itemize}

@@

We build code generators out of primitive ones using code generation 
combinators. MetaOCaml, as an instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the needed
features: to construct a code expression, to combine them, and to
execute them. The following shows the simplest code generator |one|,
and the simplest code combinators\footnote{%
$\Longrightarrow$ under an expression shows the result of its evaluation}:

\begin{code}
let one = .<1>. and plus x y = .<.~x + .~y>.
let simplest_code = let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\evalresult{.<fun x_1 -> fun y_2 -> (x_1 + (y_2 + 1))>.}
\end{code}

We use MetaOCaml brackets |.<...>.| to generate code expressions,
i.e., to construct future-stage computations. MetaOCaml provides only
one mechanism of combining code expressions, by inlining
\oleg{splicing?} one into
another. The power of that operation, called escape |.~|, comes from
the fact that the expression to be spliced in (inlined) can be
computed: escape lets us perform an arbitrary immediate code-generating
computation \emph{while} we are
building the future-stage computation. The immediate computation in
|simplest_code| is the evaluation of the function |gen|, which in turn
applies |plus|. The function |gen| receives code expressions |.<x>.|
and |.<y>.| as arguments. At the generating stage, we can manipulate
code expressions as (opaque) values. The function |gen| returns a code
expression, which is inlined in the place of the escape. MetaOCaml can
print out code expressions, so we can see the final generated code. It
has no traces of |gen| and |plus|: their applications are done at the
generation stage.

@@

To see the benefit of code generation, we notice that we can easily
parameterize our code:

\begin{code}
let simplest_param_code plus one =
  let gen x y = plus x (plus y one) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
and use it to generate code that operates on integers, floating point
numbers or booleans -- in general, any domain that implements |plus|
and |one|:
\begin{code}
let plus x y = .<.~x +. .~y>. and one = .<1.0>. in
  simplest_param_code plus one
let plus x y = .<.~x || .~y>. and one = .<true>. in
  simplest_param_code plus one
\end{code}
Running the former expression yields the function on |float|s, whereas
the latter expression is the code expression for a boolean function.
This clearly shows the separation of concerns, namely of that for domain
operations.

Let us consider a more complex expression:
\begin{code}
let param_code1 plus one =
  let gen x y = plus (plus y one) (plus x (plus y one)) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
\end{code}
with two occurrences of |plus y one|,
which may be quite a complex computation and so we would rather not do
it twice. We may be tempted to rely on the compiler's
common-subexpression elimination optimization. When the generated code is
very complex, however, the compiler may overlook common subexpressions.  Or the
subexpressions may occur in such an imperative context where the compiler
might not be able to determine if lifting them is sound. So, being
conservative, the optimizer will leave the duplicates as they are. 
We may attempt to eliminate subexpressions as follows: 
\begin{code}
let param_code1' plus one =
  let gen x y = let ce = (plus y one) in  plus ce (plus x ce) in
  .<fun x y -> .~(gen .<x>. .<y>.)>.
param_code1' plus one
\evalresult{.<fun x_1 -> fun y_2 -> ((y_2 + 1) + (x_1 + (y_2 + 1)))>.}
\end{code}
However,
the result of |param_code1' plus one| still exhibits duplicate
sub-expressions.  Our |let|-insertion optimization saved the
computation at the generating stage.  We need a combinator that
inserts the |let| expression in the generat\emph{ed} code. We need a
combinator |letgen| to be used as
\begin{code}
let ce = letgen (plus y one) in plus ce (plus x ce)
\end{code}
yielding the code like 
\begin{code}
.<let t = y + 1 in t + (x + t)>.
\end{code}
But that seems impossible because |letgen exp| has to generate
the expression |.<let t = exp in body>.| but |letgen| does not
have the |body| yet. The body needs a temporary identifier |.<t>.|
that is supposed to be the result of |letgen| itself.  Certainly
|letgen| cannot generate only part of a let-expression, without the
|body|, as all generated expressions in MetaOCaml are well-formed and
complete.

@@

\subsection{Monadic notation, making CPS code clear}\label{monadicnotation}

Comparison of the code that did let-insertion at the generating stage
\begin{code}
let ce = (plus y one) in  plus ce (plus x ce)
\end{code}
with the corresponding code inserting let at the generated code stage
\begin{code}
letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
\end{code}
clearly shows the difference between  direct-style and CPS code.
What was |let ce = init in ...| in direct style became
|init' (fun ce -> ...)| in CPS. For one thing, |let| became
``inverted''. For another, what used to be an expression that yields
a value, |init|, became an expression that takes an extra argument,
the continuation, and invokes it. The differences look negligible in
the above example. In larger expressions with many let-forms, the
number of parentheses around |fun| increases, the need to add and
then invoke the |k| continuation argument become increasingly annoying. The
inconvenience is great enough for some people to explicitly avoid CPS
or claim that numerical programmers (our users) cannot or will not
program in CPS. Clearly a better notation is needed.

@@

let l1 f = fun x -> perform \{ t <-- x; f t\}
let l2 f = fun x y -> perform \{ tx <-- x; ty <-- y; f tx ty\}

@@

Finally, |runM| runs our monad, that is, performs the computation of
the monad and returns its result, which in our case is the code
expression. We run the monad by passing it the initial state and the
initial continuation |k0|. We can now re-write our |param_code2|
example of the previous section as |param_code3|.
\begin{code}
let param_code3 plus one =
  let gen x y = bind (retN (plus y one)) (fun ce -> 
                ret (plus ce (plus x ce)))
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
\end{code}
% param_code3 plus one;;
%
That does not seem like much of an improvement. With the help of
camlp4 pre-processor, we introduce the |perform|-notation \cite{metamonadsURL},
patterned after the |do|-notation of Haskell. 
\begin{code}
let param_code4 plus one =
  let gen x y = mdo \{ ce <-- retN (plus y one);
                      ret (plus ce (plus x ce)) \}
  in .<fun x y -> .~(runM (gen .<x>. .<y>.))>.
\end{code}
The function
|param_code4|, written in the |perform|-notation, is equivalent to
|param_code3| -- in fact, the camlp4 preprocessor will convert the
former into the latter. And yet, |param_code4| looks far more
conventional, as if it were indeed in direct style.

\subsection{Generating control statements}
We can write operations that generate code other than let-statements,
e.g., conditionals: see |ifL| in Figure~\ref{ourmonad}. The function |ifL|, 
albeit straightforward, is not as general as we wish: its arguments are
already generated pieces of code rather than monadic values. We
``lift it'', see |ifM'|. 
\begin{code}
let ifM' test th el = perform \{
  testc <-- test; thc <-- th; elc <-- el;
  ifL testc thc elc\}
let gen a i = ifM' (ret .<(.~i) >= 0>.) 
                   (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 i_2 -> \protect\newline
let t_3 = (Some a_1.(i_2)) in if (i_2 >= 0) then t_3 else None>.}
let gen a i = ifM (ret .<(.~i) >= 0>.) 
                  (retN .<Some (.~a).(.~i)>.) (ret .<None>.)
 in .<fun a i -> .~(runM (gen .<a>. .<i>.))>.
\evalresult{.<fun a_1 i_2 -> 
if (i_2 >= 0) then let t_3 = (Some a_1.(i_2)) in t_3 else None>.}
\end{code}
%
We define functions |l1|,
|l2|, |l3| (analogues of |liftM|, |liftM2|, |liftM3| of Haskell) 
to make such a lifting generic. However we also need
another |ifM| function, with the same
interface (see Figure~\ref{ourmonad}).
The difference between them is
apparent: in the code above with |ifM'|, the let-insertion
happened \emph{before} the if-expression, that is, before the test that
the index |i| is positive. If |i| turned out
negative, |a.(i)| would generate an out-of-bound array access
error. On the other hand, the code with |ifM| accesses the array only
when we have verified that the index is non-negative. This example
makes it clear that the code generation (such as the one in |retN|) is 
truly an effect and we have to be clear about the sequencing of
effects when generating control constructions such as conditionals.
The form |ifM| handles such effects correctly. We
need similar operators for other OCaml control forms: for generating
case-matching statements and |for|- and |while|-loops.
\oleg{put them here?}

\section{Aspects and Functors}\label{functors}
The monad represents finer-scale code generation. We need tools for
larger-scale modularization; we can use any abstraction
mechanisms we want to structure our code generators, as long as none
of those abstractions infiltrate the generated code.

While the Object-Oriented Design community has acquired an extensive
vocabulary for describing modularity ideas, the guiding principles for
modular designs has not changed since they were first articulated by
Parnas~\cite{journals/cacm/parnas72a} and Dijkstra~
\cite{EWD:EWD447}: information hiding and separation of concerns.  To
apply these principles to the study of Gaussian Elimination, we need
to understand what are the changes between different implementations, and 
what concerns need to be addressed.  We also need to study the degree
to which these concerns are independent.
A study of Gaussian Elimination \cite{carette04} shows that
the following variations occur:
\begin{enumerate}
\vspace*{-6pt}
    \item \textbf{Domain}: In which (algebraic) domain do the
      matrix elements belong to.  Sometimes the domains are very
      specific (e.g., $\mathbb{Z}, \mathbb{Q}, \mathbb{Z}_p$ and
      floating point numbers), while in other cases the domains
      were left generic, e.g., multivariate polynomials over a
      field.  In the roughly 85 pieces of code surveyed
      \cite{carette04} 20 different domains were encountered.
    \item \textbf{Container}: Whether the matrix is represented as
      an array of arrays, a one-dimensional array, a hash table, a
      sparse matrix, etc., and whether indexing is done in C or
      Fortran style.  Additionally, if a particular representation
      had a special mechanism for efficient row exchanges.
    \item \textbf{Output choices}: Whether just the reduced
      matrix, or additionally the rank, the determinant, and the
      pivoting matrix are to be returned. In the larger algorithm
      family, routines like Maple's
      \texttt{LinearAlgebra:-LUDecomposition} have up to $2^6 +
      2^5 + 2^2 = 100$ outputs.
    \item \textbf{Fraction-free}: Whether the Gaussian Elimination
        algorithm is allowed to use unrestricted division, or only
        exact (remainder-free) division.
    \item \textbf{Pivoting}: Whether to use no, 
        column-wise, or full pivoting.
    \item \textbf{Augmented Matrices}: Whether all or only some
      columns of the matrix participate in elimination.
\end{enumerate}
\noindent In addition to the above variations, there are two aspects that 
recur frequently:
\vspace*{-6pt}
\begin{enumerate}
    \item \textbf{Length measure}:  For stability reasons
        (numerical or coefficient growth), if a domain
      possesses
        an appropriate length measure, this is sometimes used to choose
        an ``optimal'' pivot.
    \item \textbf{Normalization and zero-equivalence}: Whether the
      arithmetic operations of the domain give normalized results, and
      whether a specialized zero-equivalence routine is to be used.
\end{enumerate}
\noindent These are separated out from the others as they are cross-cutting
concerns: in the case of the length measure, a property of the domain
will influence the pivoting method \emph{if} pivoting is to be
performed.

@@

\begin{code2}
module type OUTPUT = sig
  type contr  type res
  module D : DETERMINANT   module R : RANK   module P : TRACKPIVOT
  val make_result : ('a,contr) code -> 
    (('a,res) code,
     [> `TDet of 'a D.lstate | `TRan of 'a R.lstate | `TPivot of 'a P.lstate]
       list, ('a,'w) code) monad
end
module OutDetRank(Dom:DOMAIN)(C: CONTAINER2D)
    (Det : DETERMINANT with type indet = Dom.v and type outdet = Dom.v)
    (Rank : RANK) = struct
  module Ctr = C(Dom)
  type contr = Ctr.contr
  type res = contr * Det.outdet * int
  module D = Det   module R = Rank   module P = DiscardPivot
  let make_result b = perform \{ det  <-- D.fin ();  rank <-- R.fin ();
    ret .< ( .~b, .~det, .~rank ) >. \}
end
\end{code2}

@@

\begin{figure}
\begin{code2}
module Gen(Dom: DOMAIN)(C: CONTAINER2D)(PivotF: PIVOT)
          (Update: UPDATE with type baseobj = Dom.v and type ctr = C(Dom).contr)
          (Out: OUTPUT with type contr = C(Dom).contr and type D.indet = Dom.v 
                        and type 'a D.lstate = 'a Update.D.lstate) = struct
    module Ctr = C(Dom)
    module Pivot = PivotF(Dom)(C)(Out.D)
    let gen =
      let zerobelow b r c m n brc =
        let innerbody i = perform \{
            bic <-- Ctr.get b i c;
            whenM (l1 LogicCode.not (LogicCode.equal bic Dom.zero ))
                (seqM (retLoopM (Idx.succ c) (Idx.pred m)
                          (fun k -> Update.update b r c i k) )
                      (Ctr.set b i c Dom.zero)) \} in 
        perform \{
              seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
                   (Update.update_det brc) \} in
      let dogen a = perform \{
          r <-- Out.R.decl ();
          c <-- retN (liftRef Idx.zero);
          b <-- retN (Ctr.mapper Dom.normalizerf (Ctr.copy a));
          m <-- retN (Ctr.dim1 a);
          n <-- retN (Ctr.dim2 a);
          () <-- Update.D.decl ();
          () <-- Out.P.decl ();
          seqM 
            (retWhileM (LogicCode.and_ (Idx.less (liftGet c) m)
                                       (Idx.less (liftGet r) n) )
               ( perform \{
               rr <-- retN (liftGet r);
               cc <-- retN (liftGet c);
               pivot <-- l1 retN (Pivot.findpivot b rr m cc n);
               seqM (retMatchM pivot (fun pv -> 
                        seqM (zerobelow b rr cc m n pv)
                             (Out.R.succ ()) )
                        (Update.D.zero_sign () ))
                    (Code.update c Idx.succ) \} ))
            (Out.make_result b) \} in
    .<fun a -> .~(runM (dogen .<a>.)) >.
end
\end{code2}
\end{figure}

@@

\bibliography{metamonads}
\bibliographystyle{plain}
\section{Appendix 0}
\oleg{grammar of our perform monad. Thank Lydia}
